Phylogenetics and the human microbiome.

### 2013-09-27 sbreview.tex
Alpha and beta diversity, check.

Considerable attention has been given to how the microbiome influences host phenotype.
There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
kwashiorkor \cite{smith2013gut}.

Bacteria have been the primary focus of human microbiome research, however other domains have been investigated.
Archaea and fungi interact with bacterial residents \citep{hoffmann2013archaea}.
Virus populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.

@@

As described above ``metagenome" means that data is sheared randomly.
This type of data has a complication above that taken from 16S survey data, which is that the genetic region sequenced is unknown in addition to the organism it was taken from.
However, because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.

@@

\subsection{Diversity estimates using phylogenetics}
Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature.

One approach is to apply diversity metrics to classified reads.

Here we will focus on variants of phylogenetic diversity and their use.

\citep{hill1973diversity}
Diversity and evenness: a unifying notation and its consequences

On expanded marker tree \citep{kembel2011phylogenetic}.

hill numbers
\citep{chao2010phylogenetic}

\citep{odwyer2012phylogenetic}
Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities

Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}.
We have also looked at PD under rarefaction \citep{nipperess2013mean}.

\subsection{Community comparison using phylogenetics}

The level of similarity between samples or groups thereof are often called \emph{beta diversity};
Traditional way: \citep{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale

UniFrac distance and variants has been very successful.
\citep{LozuponeKnightUniFrac05}
\citep{LozuponeEaWeightedUnifrac07}
\citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns

It recovers gradients.
\citep{nemergut2011global}
?? Global patterns in the biogeography of bacterial taxa

>>>>

Considerable attention has been given to how the microbiome influences host phenotype.
IBD, etc, NNN.
kwashiorkor \cite{smith2013gut}.
There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
Changes in archaeal and fungi have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.

@@

As described above ``metagenome" means that data is sheared randomly across the genome, thus the genetic region of a read is unknown in addition to the organism it was derived from.
Because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.

@@

\subsection{The role of Operational Taxonomic Units (OTUs)}
There continues to be a lively debate on if there is a meaningful concept of species for microbes NNN, however most microbial research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.

There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT NNN, which supplanted by UCLUST.
These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
It is well known that different ways of doing this heuristic clustering can result in very different results NNN.
There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies NNN VI-CUT.
The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

\subsection{Diversity estimates using phylogenetics}
Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature replacing taxonomic groupings with OTUs.
%The most common diversity indices are Simpson and Shannon.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.

Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities NNN \citep{odwyer2012phylogenetic}

Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
Such measures do exist \citep{chao2010phylogenetic}
We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.

\subsection{Community comparison using phylogenetics}
The level of similarity between samples or groups thereof called \emph{beta diversity}.
As with alpha diversity, it is common to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however in this case phylogenetics-based methods are widely used.
They generally go under variants of ``UniFrac'' as named by Lozupone and Knight \citep{LozuponeKnightUniFrac05;LozuponeEaWeightedUnifrac07}.
As evidence of their superiority
\citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another rather than between samples.
Weighted UniFrac is an abundance weighted version.
We have shown that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time \citep{evans2012phylogenetic}.
We also showed that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac has greater power to resolve community differences than either unweighted or weighted UniFrac.

The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
It recovers gradients.
\citep{nemergut2011global}
?? Global patterns in the biogeography of bacterial taxa
Indeed, the observation of differences between two communities is often used as prima fasce evidence of a difference between them, while the lack of such a difference indicates that the communities are not different (amish?)
